---
title: "Weekly EDA"
author: "John Foster"
date: "9/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(grid)
library(gridExtra)

# only most abundant taxa (amblyomma and ixodes)
taxa.wanted <- c("AMBAME", "AMBMAC", "AMBSP", "AMBSPP", "IXOSCA", "IXOSP", "IXOSP2", "IXOSPP", "IXOSPP1")
```

### Functions for subsetting data

The subsetting function below works as follows:

1.  Create an epiWeek column (weeks start on Sunday), which follows the CDC version of an epidemiological week. See `?lubridate::epiweek` for more detail

2. Create yearWeek column, which is just the year of observation and the epiWeek from above

3. Discard 2020 data

4. Divide that data into Training (before 2019) and Validation (2019) set

5. Then filter to nymphs of the target species

From this filtered data set, we need to find the plots where we can reasonably make forecasts. 
The first thing I did was find plots in 2019 that have at least `n.pos.obs` positive counts. 
This is `plots.2019`. 
From there, I filtered the training data set to only include the plots in `plots.2019` _and_ plots that have observations starting in 2015 (so that 2016 is a full season).
I then subset these plots further so that there are at least `n.pos.obs` positive observations in each year 2016, 2017, and 2018. 

```{r functions}
subset_targets <- function(dat.raw, spp, n.pos.obs){
  
  # dat.raw = "Exploratory_Data/data/tck_longform.Rdata"
  # spp = species to subset 
  # n.pos.obs = minimum number of positive observations in each year
  
  dat <- dat.raw %>% 
    mutate(Year = year(collectDate)) %>%    # add year column
    mutate(epiWeek = epiweek(collectDate)) %>%  # add week column, weeks start on Sunday
    mutate(yearWeek = paste(Year, epiWeek, sep = "_")) %>% # yearWeek column
    filter(Year < 2020) %>% # don't want 2020 data
    mutate(trainValidate = ifelse(Year == 2019, "Validate", "Train")) %>% # 2019 is validation
    filter(LifeStage == "Nymph") %>%  # only want nymphs
    filter(acceptedTaxonID == spp)   # only most abundant taxa (amblyomma and ixodes)

  # plots where 2019 forecasts could work
  plots.2019 <- dat %>% 
    filter(trainValidate == "Validate") %>%  # 2019 data
    filter(IndividualCount != 0) %>%
    group_by(plotID) %>% 
    count() %>%            # n = the number of positive observations in a year
    filter(n >= n.pos.obs) %>%          # want more than n.pos.obs non-zero counts in 2019
    pull(plotID) %>% 
    unique()
  
  # plots where there is long enough time series
  # and for which 2019 data exists
  plots.with.2016.obs <- dat %>% 
    filter(trainValidate == "Train") %>% 
    filter(plotID %in% plots.2019) %>% 
    group_by(plotID) %>% 
    summarise(firstObservation = min(collectDate)) %>% 
    filter(firstObservation <= ymd("2015-12-31")) %>% 
    pull(plotID)
  
  # find plots that have at least n.pos.obs positive
  # observations each year
  # need to save this object for later
  positive.obs.by.year <- dat %>% 
    filter(trainValidate == "Train") %>%
    filter(plotID %in% plots.with.2016.obs) %>% # plots that have 2016 observations
    filter(IndividualCount != 0) %>%
    group_by(plotID, Year) %>% 
    count() %>%            # n = the number of positive observations in a year
    filter(n >= n.pos.obs)         # want more than n.pos.obs non-zero counts each year
  
  # pull out possible training plots
  plots.train <- positive.obs.by.year %>% 
     pull(plotID) %>% 
     unique()       
    
  # only want the plots where there are more than n.pos.obs positive
  # counts each year - done above - and these counts occur
  # in (at least) 2016, 2017, and 2018 
  years.wanted <- 2016:2018
  plots.pos.2016.2018 <- rep(NA, length(plots.train)) # initialize 
  for(i in seq_along(plots.train)){
    # pull out years for each plot 
    subset <- positive.obs.by.year %>% 
      filter(plotID == plots.train[i]) %>% 
      pull(Year)
    
    good <- subset %in% years.wanted # match the years we want
    
    # will have same number of TRUEs if years match
    if(sum(good, na.rm = TRUE) == length(years.wanted)){
      plots.pos.2016.2018[i] <- plots.train[i] # record plotID if match
    }
  }
  # remove NAs
  plots.pos.2016.2018 <- discard(plots.pos.2016.2018, is.na)
  
  # grab good plots
  dat.all <- dat %>% 
    filter(plotID %in% plots.pos.2016.2018)
  
  return(dat.all)
}

plot_timeseries <- function(data){
    dat.plot <- data %>% 
    group_by(plotID, yearWeek) %>%  
    summarise(nymphCount = sum(IndividualCount))
  
    labels <- sort(unique(dat.plot$yearWeek))
    labels <- labels[seq(1, length(labels), length.out = 6)]
    
    ggplot(dat.plot) +
      geom_line(aes(x = (yearWeek), y = nymphCount, color = plotID, group = plotID)) +
      scale_x_discrete(breaks = labels)  
}

```

## Ixodes

```{r message=FALSE, warning=FALSE}
# load data from 01_download_clean_NEON_tick.R
dat.raw <- readRDS("Exploratory_Data/data/tck_longform.Rdata")
pos.seq <- 2:4 # number of positive observations each year
spp <- "IXOSCA"
dat <- list() # initialize

# subset to possible n.pos.obs
for(i in pos.seq){
  dat[[i]] <- subset_targets(dat.raw, spp, i)
  p <- plot_timeseries(dat[[i]]) + 
    labs(title = paste("Minimum number of positive observations each year:", i))
  print(p)
} 


```

## Amblyomma

```{r message=FALSE, warning=FALSE}
# load data from 01_download_clean_NEON_tick.R
dat.raw <- readRDS("Exploratory_Data/data/tck_longform.Rdata")
pos.seq <- 2:4 # number of positive observations each year
spp <- "AMBAME"
dat <- list() # initialize

# subset to possible n.pos.obs
for(i in pos.seq){
  dat[[i]] <- subset_targets(dat.raw, spp, i)
  p <- plot_timeseries(dat[[i]]) + 
    labs(title = paste("Minimum number of positive observations each year:", i))
  print(p)
} 



```